Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2024-09-25 manuscript/ms.tex
Update ms.tex

We now discuss in more detail incidence dispersion dynamics (distinct from but related to the dispersion parameter of the offspring distribution), how we measure them, and what they indicate.
Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious. Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution (Kendall 1954).

@@

Therefore, modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemics. A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$. The assertion that $\theta_t = I_{t-1}$ comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter. Under these conditions $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (Kendall 1954, Bjorstand 2002). However, this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another. That is, the model assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.

>>>>

We now discuss in more detail incidence dispersion dynamics (distinct from but related to the dispersion parameter of the offspring distribution), how we measure them, and what they indicate.
Infectious disease transmission follows a branching process, as new infections result from exposure to individuals who are currently infectious. Under a wide range of configurations for a branching process model, including the vast majority of parsimoneous ones, the number of infected individuals $I_t$ at time $t$ will have a negative binomial distribution \cite{kendall_stochastic_1949}.

@@

Therefore, modeling $\mu_t$ and $\theta_t$ is important to the prediction of epidemics. A time-series approximation to compartment epidemic models uses $\mu_t = \lambda I_{t-1}$ where $\lambda$ is the current local rate at which incidence at time $t-1$ produces new infections at time $t$ and $\theta_t = I_{t-1}$. The assertion that $\theta_t = I_{t-1}$ comes from the assumption that each of the individuals who acquired the infection at $t-1$ form ''independent lineages'' with identically distributed ''local'' rate parameter. Under these conditions $\theta_t = I_{t-1}$ is a good approximation for sufficiently large populations (\cite{kendall_stochastic_1949}, \cite{bjornstad_dynamics_nodate}). However, this model assumes that supopulations of infectious individuals eminating from one infected individuals are independent and statistically identical to those from another. That is, the model assumes susceptible depletion in one lineage does not affect another, transmission rates are equal across lineages, and reporting rates do not vary across lineages.
